---
title: "Homework 2"
#Author: Talea Cornelius
#Date: 9/21/2020
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r read packages}
library(tidyverse)
library(readxl)
```


Problem 0
```{r}
# •	create a public GitHub repo + local R Project; we suggest naming this repo / directory p8105_hw2_YOURUNI (e.g. p8105_hw2_ajg2202 for Jeff), but that’s not required

# •	create a single .Rmd file named p8105_hw2_YOURUNI.Rmd that renders to github_document

# •	create a subdirectory to store the local data files used in Problems 1 and 2, and use relative paths to access these data files

##.Data/Trash-Wheel-Collection-Totals-8-6-19.xlsx
##.Data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv

# •	submit a link to your repo via Courseworks

##https://github.com/taleac/p8105_hw2_tmc2184
```

Problem 1

```{r read trash data}
# Read and clean the Mr. Trash Wheel sheet:
# •	specify the sheet in the Excel file and to omit non-data entries (rows with notes / figures; columns containing notes) using arguments in read_excel
# •	use reasonable variable names
# •	omit rows that do not include dumpster-specific data
# •	round the number of sports balls to the nearest integer and converts the result to an integer variable (using as.integer)

trashwheel = read_excel("./Data/Trash-Wheel-Collection-Totals-8-6-19.xlsx", sheet=1, 
               col_types=c("numeric","text","numeric","date","numeric","numeric","numeric",
                           "numeric","numeric","numeric","numeric","numeric","numeric","numeric",
                           "skip","skip","skip"),
                           skip=1) %>%
  janitor::clean_names() %>%
  drop_na() %>%
  mutate(sports_balls = round(sports_balls,0),
         sports_balls = as.integer(sports_balls))
  
```

```{r read precipitation data}
#omit rows without precipitation data and add a variable year

#2017
precip_2017 = read_excel("./Data/Trash-Wheel-Collection-Totals-8-6-19.xlsx", sheet=6, 
               col_types=c("numeric","numeric"),
                           skip=1) %>%
  janitor::clean_names() %>%
  drop_na() %>%
  rename(precip_tot_in = total) %>%
  mutate(year = 2017)
#2018
precip_2018 = read_excel("./Data/Trash-Wheel-Collection-Totals-8-6-19.xlsx", sheet=5, 
               col_types=c("numeric","numeric"),
                           skip=1) %>%
  janitor::clean_names() %>%
  drop_na() %>%
  rename(precip_tot_in = total) %>%
  mutate(year = 2018)

```

```{r combine data}
#combine precipitation datasets and convert month to a character variable 

precip = rbind(precip_2017,precip_2018) %>%
  mutate(month = month.name[month]) %>%
  arrange(year, month)

```

```{r}
# Write a paragraph about these data; you are encouraged to use inline R. Be sure to note the number of observations in both resulting datasets, and give examples of key variables. For available data, what was the total precipitation in 2018? What was the median number of sports balls in a dumpster in 2017?
```


The dataset "Mr. Trash Wheel" includes `r ncol(trashwheel)` variables and `r nrow(trashwheel)` observations. Specific variable names are `r names(trashwheel[,1:13])` and `r names(trashwheel[,14])`.

`r names(trashwheel[,1])` is an ID variable denoting each time the dumpster is filled `r names(trashwheel[,2])` and `r names(trashwheel[,3])` specify the month and year that dumpster is filled, respectively, and `r names(trashwheel[,4])` shows the specific date. Data were collected between `r min(trashwheel[,3])` and `r max(trashwheel[,3])`. This dataset also specifies total volume of trash collected in each dumpster. Mean dumpster weight in tons over the observation period was `r round(mean(trashwheel$weight_tons),2)` and mean volume in cubic yards was `r round(mean(trashwheel$volume_cubic_yards),2)`. 

The remaining columns specify specific amounts of different types of trash, including `r names(trashwheel[,7])` (mean = `r round(mean(trashwheel$plastic_bottles),2)`, median = `r round(median(trashwheel$plastic_bottles),2)`), 
`r names(trashwheel[,8])` (mean = `r round(mean(trashwheel$polystyrene),2)`, median = `r round(median(trashwheel$polystyrene),2)`), 
`r names(trashwheel[,9])` (mean = `r round(mean(trashwheel$cigarette_butts),2)`, median = `r round(median(trashwheel$cigarette_butts),2)`), 
`r names(trashwheel[,10])` (mean = `r round(mean(trashwheel$glass_bottles),2)`, median = `r round(median(trashwheel$glass_bottles),2)`), 
`r names(trashwheel[,11])` (mean = `r round(mean(trashwheel$grocery_bags),2)`, median = `r round(median(trashwheel$grocery_bags),2)`), 
`r names(trashwheel[,12])` (mean = `r round(mean(trashwheel$chip_bags),2)`, median = `r round(median(trashwheel$chip_bags),2)`), 
and `r names(trashwheel[,13])` (mean = `r round(mean(trashwheel$sports_balls),2)`, median = `r round(median(trashwheel$sports_balls),2)`). Finally, `r names(trashwheel[,14])` shows the number of homes powered by burning the trash contained in each dumpster fill (mean = `r round(mean(trashwheel$homes_powered),2)`, median = `r round(median(trashwheel$homes_powered),2)`).


The "precipitation" dataset includes `r ncol(precip)` variables and `r nrow(precip)` observations, with `r nrow(precip)/2` observations each in 2017 and 2018. The variable `r names(precip[,1])` specifies the month, `r names(precip[,3])` specifies the year, and `r names(precip[,2])` specifies the total inches of precipitation for that given month in inches (mean = `r round(mean(precip$precip_tot_in),2)`, median = `r round(median(precip$precip_tot_in),2)`). The total precipitation in 2017 was `r aggregate(precip$precip_tot_in, by=list(Year=precip$year), FUN=sum)$x[1]` and the total precipitation in 2018 was `r aggregate(precip$precip_tot_in, by=list(Year=precip$year), FUN=sum)$x[2]`.

Problem 2
```{r read data}
# Read and clean the data; retain line, station, name, station latitude / longitude, routes served, entry, vending, entrance type, and ADA compliance. Convert the entry variable from character (YES vs NO) to a logical variable (the ifelse or recode function may be useful).

transit_data = read_csv(file = "./Data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
                       col_types = cols(
  .default = col_character(),
  `Station Latitude` = col_double(),
  `Station Longitude` = col_double(),
  `ADA` = col_logical(),
  `Free Crossover`= col_logical(),
  `Entrance Type` = col_factor(),
  `Entrance Latitude` = col_double(),
  `Entrance Longitude` = col_double()
)) %>%
  janitor::clean_names() %>%
  mutate(entry=ifelse(entry=="YES", TRUE, FALSE), route1=str_to_upper(route1)) %>%
  select(line, station_name, station_latitude, station_longitude, route1, route2, route3,
         route4, route5, route6, route7, route8, route9, route10, route11, entry, entrance_type, 
         vending, ada)
head(transit_data)
```

The "transit" dataset includes `r ncol(transit_data)` variables and `r nrow(transit_data)` observations (i.e., station entrances/exits - some stations may have multiple entrances/exits). Variable names include the line, station name, station latitude and longitude, routes, entry, entrance type, vending, and ada compliance. Thus far, I imported the data, selected the previously mentioned variables, cleaned the data names, and made sure that ada compliance and entry were coded as logical variables (i.e., true or false). These data are not tidy; route is in 11 columns (i.e., one variable spread across multiple columns).

```{r}
#•	How many distinct stations are there? Note that stations are identified both by name and by line (e.g. 125th St A/B/C/D; 125st 1; 125st 4/5); the distinct function may be useful here.
transit_data$full_name <- paste(transit_data$line, " ", transit_data$station_name)
nrow(distinct(select(transit_data, full_name, ada)))
```
There are `r nrow(distinct(select(transit_data, full_name, ada)))` distinct stations in the transit dataset.

```{r}
#•	How many stations are ADA compliant?
table(distinct(select(transit_data, full_name, ada))$ada)[2]

#•	What proportion of station entrances / exits without vending allow entrance?
novend <- subset(transit_data, vending=="NO")
table(subset(select(transit_data, entry, vending), vending == "NO")$entry)[2]/(nrow(subset(select(transit_data, entry, vending),vending == "NO")))
```
Of the `r nrow(mdat)` unique stations, `r table(distinct(select(transit_data, full_name, ada))$ada)[2]` are ADA compliant.
Of the `r nrow(novend)` station entrances/exits without vending, `r round((table(subset(select(transit_data, entry, vending), vending == "NO")$entry)[2]/(nrow(subset(select(transit_data, entry, vending),vending == "NO")))*100), 2)` percent allow entrance.

```{r restructure}
#Reformat data so that route number and route name are distinct variables. 

  lines = list("1","2","3","4","5","6","7",
                  "A","B","C","D","E","F","FS","G","GS","H","J","L","M","N","Q","R","Z")
routes=list("route1","route2","route3","route4","route5","route6","route7","route8",
            "route9","route10","route11")
    # Iterate over the list of elements one by one
    for (elem in lines)
      {for (route in routes)
    {
        if (elem %in% unlist(transit_data[route])) 
            {transit_data[elem] = TRUE}
        else
            {transit_data[elem] = FALSE}
      }   
    }
    
#drop old route variables
transit_data2 = select(transit_data, -c(route1,route2,route3,route4,route5,route6,
                                        route7,route8,route9,route10,route11))
  

#How many distinct stations serve the A train? 

#Of the stations that serve the A train, how many are ADA compliant?

```


Problem 3

```{r import pols data}
pols_data = read_csv(file = "./Data/pols-month.csv", col_types="ciiiiiiii") %>%
  janitor::clean_names() %>%
  separate(mon, into = c("year", "month", "day"), sep = "-") %>%
  mutate(month=as.integer(month),
         month = month.name[month], 
         president=prez_gop) %>%
  mutate(president=recode(president, `0` = "dem", `1` = "gop", `2` = "gop")) %>%
  select(-c(prez_dem,prez_gop,day))
```

```{r import snp data}
snp_data = read_csv(file = "./Data/snp.csv", col_types="cd") %>%
  janitor::clean_names() %>%
  separate(date, into = c("month", "day", "year"), sep = "/")%>%
  mutate(month = month.name[month]) %>%
  select(year, month,close)

```



Third, tidy the unemployment data so that it can be merged with the previous datasets. This process will involve switching from “wide” to “long” format; ensuring that key variables have the same name; and ensuring that key variables take the same values.
Join the datasets by merging snp into pols, and merging unemployment into the result.
Write a short paragraph about these datasets. Explain briefly what each dataset contained, and describe the resulting dataset (e.g. give the dimension, range of years, and names of key variables).
Note: we could have used a date variable as a key instead of creating year and month keys; doing so would help with some kinds of plotting, and be a more accurate representation of the data. Date formats are tricky, though. For more information check out the lubridate package in the tidyverse.



