---
title: "Homework 2"
#Author: Talea Cornelius
#Date: 9/21/2020
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r read packages}
library(tidyverse)
library(readxl)
```


Problem 0
```{r}
# •	create a public GitHub repo + local R Project; we suggest naming this repo / directory p8105_hw2_YOURUNI (e.g. p8105_hw2_ajg2202 for Jeff), but that’s not required

# •	create a single .Rmd file named p8105_hw2_YOURUNI.Rmd that renders to github_document

# •	create a subdirectory to store the local data files used in Problems 1 and 2, and use relative paths to access these data files

##.Data/Trash-Wheel-Collection-Totals-8-6-19.xlsx
##.Data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv

# •	submit a link to your repo via Courseworks

##https://github.com/taleac/p8105_hw2_tmc2184
```

Problem 1

```{r read trash data}
# Read and clean the Mr. Trash Wheel sheet:
# •	specify the sheet in the Excel file and to omit non-data entries (rows with notes / figures; columns containing notes) using arguments in read_excel
# •	use reasonable variable names
# •	omit rows that do not include dumpster-specific data
# •	round the number of sports balls to the nearest integer and converts the result to an integer variable (using as.integer)

trashwheel = read_excel("./Data/Trash-Wheel-Collection-Totals-8-6-19.xlsx", sheet=1, 
               col_types=c("numeric","text","numeric","date","numeric","numeric","numeric",
                           "numeric","numeric","numeric","numeric","numeric","numeric","numeric",
                           "skip","skip","skip"),
                           skip=1) %>%
  janitor::clean_names() %>%
  drop_na() %>%
  mutate(sports_balls = round(sports_balls,0),
         sports_balls = as.integer(sports_balls))
  
```

```{r read precipitation data}
#omit rows without precipitation data and add a variable year

#2017
precip_2017 = read_excel("./Data/Trash-Wheel-Collection-Totals-8-6-19.xlsx", sheet=6, 
               col_types=c("numeric","numeric"),
                           skip=1) %>%
  janitor::clean_names() %>%
  drop_na() %>%
  rename(precip_tot_in = total) %>%
  mutate(year = 2017)
#2018
precip_2018 = read_excel("./Data/Trash-Wheel-Collection-Totals-8-6-19.xlsx", sheet=5, 
               col_types=c("numeric","numeric"),
                           skip=1) %>%
  janitor::clean_names() %>%
  drop_na() %>%
  rename(precip_tot_in = total) %>%
  mutate(year = 2018)

```

```{r combine data}
#combine precipitation datasets and convert month to a character variable 

precip = rbind(precip_2017,precip_2018) %>%
  mutate(month = month.name[month]) %>%
  arrange(year, month)

```

```{r}
# Write a paragraph about these data; you are encouraged to use inline R. Be sure to note the number of observations in both resulting datasets, and give examples of key variables. For available data, what was the total precipitation in 2018? What was the median number of sports balls in a dumpster in 2017?
```


The dataset "Mr. Trash Wheel" includes `r ncol(trashwheel)` variables and `r nrow(trashwheel)` observations. Specific variable names are `r names(trashwheel[,1:13])` and `r names(trashwheel[,14])`.

`r names(trashwheel[,1])` is an ID variable denoting each time the dumpster is filled `r names(trashwheel[,2])` and `r names(trashwheel[,3])` specify the month and year that dumpster is filled, respectively, and `r names(trashwheel[,4])` shows the specific date. Data were collected between `r min(trashwheel[,3])` and `r max(trashwheel[,3])`. This dataset also specifies total volume of trash collected in each dumpster. Mean dumpster weight in tons over the observation period was `r round(mean(trashwheel$weight_tons),2)` and mean volume in cubic yards was `r round(mean(trashwheel$volume_cubic_yards),2)`. 

The remaining columns specify specific amounts of different types of trash, including `r names(trashwheel[,7])` (mean = `r round(mean(trashwheel$plastic_bottles),2)`, median = `r round(median(trashwheel$plastic_bottles),2)`), 
`r names(trashwheel[,8])` (mean = `r round(mean(trashwheel$polystyrene),2)`, median = `r round(median(trashwheel$polystyrene),2)`), 
`r names(trashwheel[,9])` (mean = `r round(mean(trashwheel$cigarette_butts),2)`, median = `r round(median(trashwheel$cigarette_butts),2)`), 
`r names(trashwheel[,10])` (mean = `r round(mean(trashwheel$glass_bottles),2)`, median = `r round(median(trashwheel$glass_bottles),2)`), 
`r names(trashwheel[,11])` (mean = `r round(mean(trashwheel$grocery_bags),2)`, median = `r round(median(trashwheel$grocery_bags),2)`), 
`r names(trashwheel[,12])` (mean = `r round(mean(trashwheel$chip_bags),2)`, median = `r round(median(trashwheel$chip_bags),2)`), 
and `r names(trashwheel[,13])` (mean = `r round(mean(trashwheel$sports_balls),2)`, median = `r round(median(trashwheel$sports_balls),2)`). Finally, `r names(trashwheel[,14])` shows the number of homes powered by burning the trash contained in each dumpster fill (mean = `r round(mean(trashwheel$homes_powered),2)`, median = `r round(median(trashwheel$homes_powered),2)`).


The "precipitation" dataset includes `r ncol(precip)` variables and `r nrow(precip)` observations, with `r nrow(precip)/2` observations each in 2017 and 2018. The variable `r names(precip[,1])` specifies the month, `r names(precip[,3])` specifies the year, and `r names(precip[,2])` specifies the total inches of precipitation for that given month in inches (mean = `r round(mean(precip$precip_tot_in),2)`, median = `r round(median(precip$precip_tot_in),2)`). The total precipitation in 2017 was `r aggregate(precip$precip_tot_in, by=list(Year=precip$year), FUN=sum)$x[1]` and the total precipitation in 2018 was `r aggregate(precip$precip_tot_in, by=list(Year=precip$year), FUN=sum)$x[2]`.

Problem 2
```{r read data}
# Read and clean the data; retain line, station, name, station latitude / longitude, routes served, entry, vending, entrance type, and ADA compliance. Convert the entry variable from character (YES vs NO) to a logical variable (the ifelse or recode function may be useful).

```


Write a short paragraph about this dataset – explain briefly what variables the dataset contains, describe your data cleaning steps so far, and give the dimension (rows x columns) of the resulting dataset. Are these data tidy?
Answer the following questions using these data:
•	How many distinct stations are there? Note that stations are identified both by name and by line (e.g. 125th St A/B/C/D; 125st 1; 125st 4/5); the distinct function may be useful here.
•	How many stations are ADA compliant?
•	What proportion of station entrances / exits without vending allow entrance?
Reformat data so that route number and route name are distinct variables. How many distinct stations serve the A train? Of the stations that serve the A train, how many are ADA compliant?



Problem 3
This problem uses the FiveThirtyEight data; these data were gathered to create the interactive graphic on this page. In particular, we’ll use the data in pols-month.csv, unemployment.csv, and snp.csv. Our goal is to merge these into a single data frame using year and month as keys across datasets.
First, clean the data in pols-month.csv. Use separate() to break up the variable mon into integer variables year, month, and day; replace month number with month name; create a president variable taking values gop and dem, and remove prez_dem and prez_gop; and remove the day variable.
Second, clean the data in snp.csv using a similar process to the above. For consistency across datasets, arrange according to year and month, and organize so that year and month are the leading columns.
Third, tidy the unemployment data so that it can be merged with the previous datasets. This process will involve switching from “wide” to “long” format; ensuring that key variables have the same name; and ensuring that key variables take the same values.
Join the datasets by merging snp into pols, and merging unemployment into the result.
Write a short paragraph about these datasets. Explain briefly what each dataset contained, and describe the resulting dataset (e.g. give the dimension, range of years, and names of key variables).
Note: we could have used a date variable as a key instead of creating year and month keys; doing so would help with some kinds of plotting, and be a more accurate representation of the data. Date formats are tricky, though. For more information check out the lubridate package in the tidyverse.



